Cheminformatics: The Computer Science of Chemical Discovery

### 2011-08-15 paper.tex

\additionalauthors{Additional authors:
Andreas Bender (University of Cambridge, email: {\texttt{bender.andreas@gmail.com}}),
Jean-Loup Faulon (University of Evry, email: {\texttt{Jean-Loup.Faulon@issb.genopole.fr}}),
Janna Hastings (European Bioinformatics Institute, Cambridge, UK, email: {\texttt{janna.hastings@gmail.com}}),
Noel O'Boyle (University College Cork, Cork, Ireland, email: {\texttt{baoilleach@gmail.com}}),
John Overington (European Bioinformatics Institute, Cambridge, UK, email: {\texttt{jpo@ebi.ac.uk}}),
Herman Van Vlijmen (Tibotec, Beerse, Belgium, email: {\texttt{hvvlijme@its.jnj.com}}), and
Egon Willighagen (Karolinska Institutet, Stockholm, Sweden, email: {\texttt{egon.willighagen@ki.se}})
.}
\date{25 June 2011}

\maketitle
\begin{abstract}
One of the most prominent success stories in all the sciences over the last decade has
  been the advance of bioinformatics: interdisciplinary collaboration between
 computer scientists and molecular biologists led to the sequencing of the human genome
  and many other accomplishments, since researchers connected biological
  concepts to the theory of string algorithms. However, despite this great
  success, few computer scientists are familiar with a related (and older!)
  discipline---cheminformatics, the use of computers to represent
 the structures of small molecules and analyze their properties.
  Cheminformatics has a wide applicability,
  from the discovery of novel drug candidates to the optimization of
  certain physicochemical properties of small molecules. Until
 recently, the data and techniques of cheminformatics have been closely guarded
  secrets of companies whose financial success depended on being the first to produce
  the new ``miracle molecule.'' Only within the last decade---and largely because of
 chemists volunteering their time for an Open Science ``movement''---have researchers
  gained access to freely available software packages and databases of tens of
  millions of chemicals. Academic chemists now confront a variety of unsolved
  algorithmic problems that could not have been tackled a decade ago, but whose
  solutions are critical to research ranging from determining the behavior of small
  molecules in biological pathways, to finding a cure for malaria.
\end{abstract}

\category{H.4}{Information Systems Applications}{Miscellaneous}
\terms{
\begin{itemize}
\item \textit{Bioassay} - A system measuring the biological activity of a chemical compound on a particular biological assay (with all its parameters).
\item \textit{Biodegradation} - is the chemical dissolution of materials by bacteria or other biological means.
\item \textit{Conformation and conformational isomerism} - In chemistry, conformational isomerism is a form of stereoisomerism in which the isomers can be interconverted exclusively by rotations about formally single bonds. Single possible states of compounds are called conformers.
\item \textit{Ligand (biochemistry)} - a substance that binds to a protein.
\item \textit{Ligand (coordination chemistry)} - an atom, ion, or functional group that donates one or more of its electrons through a coordinate covalent bond to one or more central atoms or ions.
\item \textit{Mass spectrometry} - Mass spectrometry (MS) is an analytical technique that measures the mass-to-charge ratio of charged particles. It is used for determining masses of particles, for determining the elemental composition of a sample or molecule, and for elucidating the chemical structures of molecules, such as peptides and other chemical compounds.
\item \textit{Metabolites} - Metabolites are the intermediates and products of metabolism. The term metabolite is usually restricted to small molecules (chemical compounds).
\item \textit{Proteochemometric modeling} - combining data mining for chemical compounds (ligands) and proteins within the same machine learning framework.
\item \textit{Protonation and protonation state} - In chemistry, protonation is the addition of a proton (H$^+$) to an atom, molecule, or ion. The different states (different number of H$^+$ or on different atoms) are called protonation states for a molecular compound (e.g. de-protonated, mono-protonated, di-protonated, or any variations of those states).
\item \textit{Reaction catalysis} - Catalysis is the change in rate of a chemical reaction due to the participation of a substance called a catalyst.
\item \textit{Receptor (biochemistry)} - in biochemistry, a protein molecule that receives and responds to a neurotransmitter, or other substance.
\item \textit{Regulation authorities} - In the US the Food and Drug Administration (FDA) is responsible for protecting and promoting public health through the regulation and supervision of e.g. dietary supplements, prescription and over-the-counter pharmaceutical drugs (medications), vaccines, biopharmaceuticals, medical devices, and many other health related items. The European Medicines Agency (EMA) is a European agency for the evaluation of medicinal products.
\item \textit{SMARTS} - SMiles ARbitrary Target Specification \\(SMARTS) is a language for specifying substructure patterns in molecules
\item \textit{Stereoisomer/stereoisomerism} - Stereoisomers are isomeric molecules that have the same molecular formula and sequence of bonded atoms (constitution), but that differ only in the three-dimensional orientations of their atoms in space.
\item \textit{Tautomerism} - Tautomers are isomers (same molecular formula but different structural formula) of organic compounds that readily interconvert by a chemical reaction called tautomerization.
\end{itemize}
}
\keywords{cheminformatics, graphs, chemistry, databases, knowledge management, open source, open data}

@@

Novel technologies in the life sciences churn out information at an ever
increasing rate---public data stores such as the one at the European Bioinformatics
Institute (EBI) contain on the order of 10 petabytes of biological information.
However, while biological information has been publicly available for decades,
the same has not been true for chemical information until very recently. It was
not until 2004 that a large public small molecule structure repository (PubChem)
became freely available. This was followed by other databases, as discussed later
in this article. In the same vein, open source software and the publication of
algorithms have only recently become a main focus of the field
\cite{faulon2010}.

@@

Making chemical information public is important because the
drugs we take are the result of intensive and costly research,
which is facilitated by the extent of information available. The
more information is shared (a process very hard to achieve
in pharmaceutical companies), the more information is available to each
researcher, facilitating the development of
novel treatments. The relevance of computer
science in this research becomes clear in considering the quantity of data available
currently---74 million different molecules are currently stored in
public databases (such as PubChem), together with an even larger number of annotations.
The design of data structures for this chemical information, as well as subsequent
data mining of this information, are areas where expertise in computer
science is urgently needed. (For a recent, detailed introduction to
cheminformatics, written for computer scientists, see
\cite{brown2009}).

Cheminformatics comprises different areas, which broadly speaking can be divided
into three fields: \emph{capturing data} (using lab notebooks or potentially using
formats such as the Chemical Markup Language for publications); \emph{storing data}
(designing database schemas, devising ontologies) and \emph{mining data} (such as for
bioactivity prediction, which might involve algorithms based on graphs).
However, a major difference from bioinformatics lies in the type of
information being analyzed: While bioinformatics often deals with sequences, the
domain of cheminformatics is chemical structures. In the case of the former,
information can very frequently be represented as one-dimensional strings, which
are relatively easy to handle computationally. In the case of the latter, chemical
structures may possess rings, branches as well as multiple valid representations
of the same structure, potentially giving rise to ambiguities. Hence,
chemical structure is often reported to be more difficult to standardize.

@@

While cheminformatics has a broad range of applicability, ranging from
agrochemical research and drug discovery to the design of novel
materials, we have chosen to use drug discovery as the context of this
article. The remainder of this article is structured around the
concept of ``risk minimization'' in drug discovery---how does one
minimize the chances of a small molecule failing as a drug candidate
and how do cheminformatics methods play a key role in these efforts.

@@

Most cheminformatics applications rely on large databases of chemical structures, 
their properties, and relationships to e.g. biological targets.
Organizing and maintaining such databases, searching based
on features of chemical structures, and clustering similar structures together,
are essential utilities to enable many scientific applications. Yet, each one of
those areas poses its unique computer science challenges.

There is considerable progress in open chemical and bioactivity databases like
ChEMBL and NCBI-PubChem, which are both freely available. Still, the integration
of chemical databases with each other remains a challenge, not only due to the
normalization problems in the chemical and bioactivity landscape, but also due
to sheer data volume. As mentioned earlier, we have to be able to
search in complex data, for example multi-labelled molecular graphs, where
graph labels can be different. Still, we need to be able to capture
molecular graphs in a risk-bounded machine encoded way.

Chemical graphs can differ based on the manner in which the structures
are drawn. Also chemical behaviors can result in multiple encodings of
``the same'' molecule and its tautomeric states, different protonation
states, or many variations of implicit chemical expert
knowledge. Checking whether two graphs represent the same chemical is
an expensive operation (the next section will discuss this in-depth);
thus, a representation of the structure that is unique and invariant
with respect to how the structure is drawn is needed for
identification. An early effort in this direction was the Simplified
Molecular Input Line Specification (SMILES), and specifically the
generation of a single canonical SMILES for a particular structure. However,
over the course of time, several
different SMILES implementations became available, with no universal
standard for generating a canonical SMILES.
%After years of research and various
%commercial products, there is today no single public standard for a
%line-notation of molecular labeled graphs, which is available to a
%broad audience (academic and
%commercial). %The very same problem does also exist for
%bioassay ontologies and the lack of options to normalize the property spaces of
%chemical compounds, e.g. biological assays or clinical data outcome
%(biobanking).

With each database using their own algorithm for molecular encoding,
automated exchange of data between different databases was hindered.
As more and more data became available online,
structure-based identification became a pressing need, and the IUPAC
International Chemical Identifier (InChI) code was developed. The
InChI is a non-proprietary, structured textual identifier for chemical
entities \cite{inchi}. InChI identifiers are not intended to be read and
understood by humans, but are useful for computational matching of
chemical entities. For rapid database lookups, the InChIKey is a
hashed key for the InChI, which has an invariant length of 14
characters. Figure \ref{smiles} illustrates the SMILES, InChI and InChIKey for
lipoic acid.

@@

InChI or other identity-mapping algorithms allow for exact searching (within the
limitations of the algorithms), but two other requirements for scientific
discovery based on chemical databases are \emph{substructure searching} and \emph{similarity
searching}. In substructure searching, the database is searched for a specified
wholly contained part of the structures in the database; in similarity
searching, the database is searched for structures similar to a provided search
structure. Chemical search packages are often implemented and optimized for a
given database technology, for example, the OrChem package is an open source
chemical search package for the Oracle database \cite{rijnbeek2009}.

Graph substructure matching is a variant of \emph{graph isomorphism detection},which is
widely believed to be computationally intractable [3]. To execute a graph isomorphism
search across a full chemical database of thousands or even millions of
structures is infeasible. For that reason, complex heuristics must be
used to drastically filter the candidate search space. Structure-based
\emph{fingerprints} are commonly used for this purpose. Fingerprints encode
characteristic features of a given chemical structure in a Boolean array, or
bitmap. The fingerprint is created by an algorithm which generates bitmap
patterns for each atom in the structure; then each atom and its nearest
neighbors, including the bonds between them; then each group of atoms connected
by paths of increasing lengths up to the maximum implemented for the algorithm
(commonly around eight). The bitmap patterns are added together into the final
fingerprint. If all bits in a query fingerprint are also present in the target
fingerprint of a stored database structure, only then is the structure subjected
to the computationally expensive subgraph matching algorithm. Bit operations are
very fast and independent of the number of atoms in a structure due to the fixed
length of the fingerprint.

Fingerprints also facilitate the rapid calculation of quantitative
structure-based similarity measures. An example measure of similarity is the
\emph{Tanimoto coefficient}, calculated as the ratio $T(a,b) = \frac{c}{(a+b-c)}$,
where $c$ is the count of bits on in the same position in both the two
fingerprints, $a$ is the count of bits on in the first structure, and $b$ is the
count of bits on in the second structure. The Tanimoto coefficient varies in the
range 0.0 -- 1.0, with a score of 1.0 indicating that the two structures are very
similar (i.e. their fingerprints are the same). Structure-based similarity is of
limited use in grouping of chemicals based on non-structural features, such as
shared bioactivity profiles. For this purpose, a large-scale and flexible
classification structure is needed for chemicals and their linked biological
information. One such answer to this need is provided in the form of \emph{ontologies}.
An ontology is a formal specification of entities and their relationships in a
domain of interest. It is structured around an underlying logical representation
such as Description Logics~\cite{baaderdl2007}. The logic-based representation allows
automated reasoning to derive new knowledge (inferences) from the knowledge
encoded, making the representation both compact and powerful.
Ontologies are already in widespread use across the biomedical domain. A
similarity measure based on the information encoded in ontologies is termed
\emph{semantic similarity}, and such measures have already been applied to enhance
chemical classification in several problem areas \cite{couto2010}. Many open
questions still remain about the best way to combine the information encoded in
chemical structures with chemical ontologies \cite{hastingsowled2010}.

@@

Enumerating molecules is a combinatorial problem that has fascinated
chemists, computer scientists and mathematicians alike for more than a
century. It is one important risk-reduction element in still being
able to encode closely-related molecules and for keeping the space
complexity of databases under (algorithmic) control. Indeed, while
trying to solve the problem of counting the isomers of paraffin
structures or counting substituted aromatic compounds, fundamental
principles in graph theory and combinatorics were developed by Cayley,
Polya and others. Even terms that are widely-used nowadays like graph
and tree were originally coined in a chemistry context (Chapters 1 and
8 in \cite{faulon2010}). More than four decades ago, Joshua Lederberg,
Carl Djerassi, Edward Feigenbaum and others from Stanford University
started developing codes to explicitly enumerate molecules. Their
project, named DENDRAL, was funded by NASA with the goal of developing
an automated system to elucidate compounds from spectral data. While
studying this challenging problem, important concepts in computer
science were developed. DENDRAL is generally quoted in artificial
intelligence textbooks as a pioneer project that produced the first
expert system.

Enumerating molecules is not only an interesting academic exercise but has
practical utility as well. The foremost application of enumeration is \emph{structure
elucidation} of natural compounds such as metabolites.
Ideally, the wishful bench chemist collects experimental data
from an unknown compound, the data is fed into a code, and the resulting
structure is returned. Although that streamlined process may not yield a unique
solution, there are commercial software products such as MOLGEN
(http://molgen.de) that can, for instance, list all structures matching a given
molecular formula, or mass spectrometry (MS) data. Another important application
of enumeration is \emph{molecular design}. Here the problem is to design compounds
(drugs, for instance) that optimize some physical, chemical, or biological
property or activity. Although less prolific and more recent than structure
elucidation, molecular design has introduced some novel solutions to molecular
enumeration. Finally, with the advent of combinatorial chemistry, molecular
enumeration has taken a central role as it allows computational chemists to
construct virtual libraries and test hypotheses, and it provides guidance in the
design of optimal combinatorial experiments.

The major difficulty with enumeration is that the \emph{in silico}
representation of molecules, the so-called molecular graphs, are
labeled objects (i.e. labeled graphs), while the atoms of a molecule
are of course not uniquely labeled. The mathematical concept to tackle
this problem is to consider orbits of labeled molecular graphs under
the operation of the symmetric group. When enumerating molecular
structures, one has thus to solve the so--called isomorphism problem.
The good news is that isomorphism can be solved in polynomial time for
molecular graphs due to the fact that molecules belong to a restricted
class of graphs known as \emph{bounded valence graphs}; however, that
same restriction prevents applying to a molecular graph the polynomial
delay enumeration results obtained for general graphs (Chapter 8 in
\cite{faulon2010}). While the complexity of molecular graph
enumeration remains an open problem it has nonetheless been shown that
molecules can be sampled efficiently \cite{goldberg1999}. Sampling
procedures based on the Metropolis or the Genetic Algorithms have been
developed and used to elucidate natural compounds from NMR data.

The underlying computational complexity and potential intractability
of molecular enumeration may not be a bottleneck after all, as
computational chemists have developed and successfully used
enumeration tools to generate large chemical libraries such as GDB-13,
comprising almost a trillion chemical
structures\footnote{http://www.dcb-server.unibe.ch/groups/reymond/gdb/home.html}. Yet,
the current enumeration software products do not generally produce
stereoisomers or tautomers; these require specific enumeration
procedures, which are still the subject of investigations by the
cheminformatics community.

@@

The basis of predictive modeling in cheminformatics is that
\emph{chemical structure implies biological activity}. Thus, the goal
of any modeling approach is to capture and characterize the
correlations between structural features and the observed biological
activities. The goal is to describe cause-effect relationships with
the lowest possible risk, and also to describe the likelihood of error
(confidence) when using such models for decision making. In many
cases, the activity of a small molecule is due to its interaction with
a receptor. Traditional QSAR approaches ignore receptor interactions,
focusing only on small molecule features, and therefore loose valuable
information. As a result, techniques such as proteochemometric
modeling have been designed to take into account both ligand and
receptor structures.

>>>>

\additionalauthors{Additional authors:
Andreas Bender (University of Cambridge, email: {\texttt{andreas.bender@cantab.net}}),
Jean-Loup Faulon (University of Evry, email: {\texttt{Jean-Loup.Faulon@issb.genopole.fr}}),
Janna Hastings (European Bioinformatics Institute, Cambridge, UK, email: {\texttt{janna.hastings@gmail.com}}),
Noel O'Boyle (University College Cork, Cork, Ireland, email: {\texttt{baoilleach@gmail.com}}),
John Overington (European Bioinformatics Institute, Cambridge, UK, email: {\texttt{jpo@ebi.ac.uk}}),
Herman Van Vlijmen (Tibotec, Beerse, Belgium, email: {\texttt{hvvlijme@its.jnj.com}}), and
Egon Willighagen (Karolinska Institutet, Stockholm, Sweden, email: {\texttt{egon.willighagen@ki.se}})
.}
\date{25 June 2011}

\maketitle
\begin{abstract}
One of the most prominent success stories in all the sciences over the last decade has
  been the advance of bioinformatics: The interdisciplinary collaboration between
 computer scientists and molecular biologists led as one its major accomplishments to the sequencing of the human genome, where researchers connected biological concepts to the theory of string algorithms. However, despite this great
  success, few computer scientists are familiar with a related (and older!)
  discipline: cheminformatics, the use of computers to represent
 the structures of small molecules and analyze their properties.
  Cheminformatics has wide applicability, from the discovery of novel drug candidates to the optimization of
  certain physicochemical properties of small molecules. Until
 recently, the data and techniques employed in the cheminformatics field have been closely guarded
  secrets of companies whose financial success depended on being the first to produce
  the new ``miracle molecule.'' Only within the last decade---and as an effect of a change in mindsets, government-mandated policy changes and last but not least because of  chemists volunteering their time for an Open Science ``movement''---have researchers gained access to freely available software packages and databases of tens of
  millions of chemicals. Academic chemists now confront a variety of unsolved
  algorithmic problems that could not have been tackled a decade ago, but whose
  solutions are critical to research ranging from determining the behavior of small
  molecules in biological pathways, to finding a cure for malaria as well as many other neglected diseases.
\end{abstract}

\category{H.4}{Information Systems Applications}{Miscellaneous}
\terms{
\begin{itemize}
\item \textit{Bioassay} - A system measuring the biological activity of a chemical compound on a particular biological assay (with all its parameters).
\item \textit{Biodegradation} - is the chemical dissolution of materials by bacteria or other biological means.
\item \textit{Conformation and conformational isomerism} - In chemistry, conformational isomerism is a form of stereoisomerism in which the isomers can be interconverted exclusively by rotations about formally single bonds. Single possible states of compounds are called conformers.
\item \textit{Ligand (biochemistry)} - a substance that binds to a protein.
\item \textit{Ligand (coordination chemistry)} - an atom, ion, or functional group that donates one or more of its electrons through a coordinate covalent bond to one or more central atoms or ions.
\item \textit{Mass spectrometry} - Mass spectrometry (MS) is an analytical technique that measures the mass-to-charge ratio of charged particles. It is used for determining masses of particles, for determining the elemental composition of a sample or molecule, and for elucidating the chemical structures of molecules, such as peptides and other chemical compounds.
\item \textit{Metabolites} - Metabolites are the intermediates and products of metabolism. The term metabolite is usually restricted to small molecules (chemical compounds).
\item \textit{Proteochemometric modeling} - combining data mining for chemical compounds (ligands) and proteins within the same machine learning framework.
\item \textit{Protonation and protonation state} - In chemistry, protonation is the addition of a proton (H$^+$) to an atom, molecule, or ion. The different states (different number of H$^+$ or on different atoms) are called protonation states for a molecular compound (e.g. de-protonated, mono-protonated, di-protonated, or any variations of those states).
\item \textit{QSAR} - Quantitative Structure-Activity Relationships; methods to relate the structure and the biological activity of a chemical structure in a quantitative manner.
\item \textit{Reaction catalysis} - Catalysis is the change in rate of a chemical reaction due to the participation of a substance called a catalyst.
\item \textit{Receptor (biochemistry)} - in biochemistry, a protein molecule that receives and responds to a neurotransmitter, or other substance.
\item \textit{Regulation authorities} - In the US the Food and Drug Administration (FDA) is responsible for protecting and promoting public health through the regulation and supervision of e.g. dietary supplements, prescription and over-the-counter pharmaceutical drugs (medications), vaccines, biopharmaceuticals, medical devices, and many other health related items. The European Medicines Agency (EMA) is a European agency for the evaluation of medicinal products.
\item \textit{SMARTS} - SMiles ARbitrary Target Specification \\(SMARTS) is a language for specifying substructure patterns in molecules
\item \textit{Stereoisomer/stereoisomerism} - Stereoisomers are isomeric molecules that have the same molecular formula and sequence of bonded atoms (constitution), but that differ only in the three-dimensional orientations of their atoms in space.
\item \textit{Tautomerism} - Tautomers are isomers (same molecular formula but different structural formula) of organic compounds that readily interconvert by a chemical reaction called tautomerization.
\end{itemize}
}
\keywords{cheminformatics, chemoinformatics, graphs, chemistry, databases, knowledge management, open source, open data}

@@

Novel technologies in the life sciences churn out information at an ever
increasing rate---public data stores such as the one at the European Bioinformatics
Institute (EBI) contain on the order of 10 petabytes of biological information.
However, while biological information has been publicly available for decades,
the same has not been true for chemical information for very long. It was
not until 2004 that a large public small molecule structure repository (PubChem)
became freely available. This was followed by other databases, as discussed later
in this article. In the same vein, open source software and the publication of
algorithms have only recently become a main focus of the field
\cite{faulon2010}.

@@

Making chemical information public is important because the
drugs we take are the result of intensive and costly research,
which is facilitated by the extent of information available. The
more information is shared (a process very hard to achieve
in pharmaceutical companies), the more information is available to each
researcher and facilitating the development of
novel treatments. The relevance of computer
science in this research becomes clear in considering the quantity of data available---
currently 74 million different molecules are stored in
public databases (such as PubChem), together with an even larger number of annotations.
The design of data structures for this chemical information, as well as subsequent
data mining of this information, are areas where expertise in computer
science is urgently needed. (For a recent, detailed introduction to
cheminformatics, written for computer scientists, see
\cite{brown2009}).

Cheminformatics comprises different areas, which broadly speaking can be divided
into three fields: \emph{capturing data} (using lab notebooks or potentially using
formats such as the Chemical Markup Language for publications); \emph{storing data}
(designing database schemas, devising ontologies) and \emph{mining data} (such as for
bioactivity prediction, which might involve algorithms based on graphs).
However, a major difference to the area of bioinformatics lies in the type of
information being analyzed: While bioinformatics often deals with sequences, the
domain of cheminformatics is chemical structures. In the case of the former,
information can very frequently be represented as one-dimensional strings, which
are relatively easy to handle computationally. In the case of the latter however, chemical
structures may possess rings, branches as well as multiple valid representations
of the same structure (such as so-called 'tautomers', where hydrogen atoms can be positioned in different places of the molecules), potentially giving rise to ambiguities. Hence,
chemical structures are often reported to be more difficult to standardize.

@@

While cheminformatics has a rather broad applicability, ranging from
agrochemical research and drug discovery to the design of novel
materials (since everything around us consists of 'molecules'), we have chosen to use drug discovery as the context of this
article due to its central relevance for human wellbeing. Given that the safety of drugs on the market is a key concern both for the general audience as well as researchers in pharmaceutical companies, the remainder of this article is structured around the
concept of ``risk minimization'' in drug discovery---the question of how to minimize the chances of a small molecule failing in the various research and development stages as a drug candidate, and how to successfully employed cheminformatics methods in these efforts.

@@

Most cheminformatics applications rely on large databases of chemical structures, 
their properties, and relationships to e.g. biological targets.
Organizing and maintaining such databases, as well as searching and clustering similar structures together,
are essential  to enable many scientific applications. Yet, each one of
those areas poses its unique computer science challenges.

There is considerable progress in open chemical and bioactivity databases like
ChEMBL and NCBI-PubChem, which are both freely available and contain in the order of millions (ChEMBL) and tens of millions of data points each (PubChem). Still, the integration
of chemical databases with each other remains a challenge, not only due to the
normalization problems in the chemical and bioactivity landscape, but also due
to sheer data volume. One problem is the annotation with the 'protein targets' a molecule binds to - here dozens of protein and gene identifiers exist, generally without the availability of a one-to-one mapping from each ontology, which makes merging databases often a very cumbersome exercise.

Given the requirement to 'make sense' of as much chemical data at hand as possible, we have to be able to search in complex data, for example multi-labelled molecular graphs, where graph labels can be different. Hence, we need to be able to capture
molecular graphs in a risk-bounded machine encoded way.

One main challenge in the cheminformatics field is that not one unique way exists to represent every molecule in a machine-readable form. This results in multiple encodings of
``the same'' molecule and its different tautomeric or protonation states (see glossary for an explanation of the technical terms). Checking whether two graphs represent the same chemical structure is
an expensive operation (the next section will discuss this in-depth);
thus, a representation of the structure that is unique and invariant
with respect to a particular graphical display of that chemical is needed for
identification. An early effort in this direction was the Simplified
Molecular Input Line Specification (SMILES), and specifically the
generation of a single, 'canonical' SMILES for a particular structure. The idea was to devise rules for a unique ordering of the atoms in a molecular structure, hence resulting in the same representation of the same molecule. However,
over the course of time, several
different SMILES implementations became available, with no universal
standard for generating a canonical SMILES. Nonetheless, the SMILES format remains one of the major representations of molecules used in practice today.

With each database using their own algorithm for molecular encoding,
automated exchange of data between different databases was hindered.
As more and more data became available online, the use of unique
structure-based identifiers became a pressing need, resulting in the IUPAC
International Chemical Identifier (InChI) to be developed. The
InChI is a non-proprietary, structured textual identifier for chemical
entities \cite{inchi}. InChI identifiers are not intended to be read and
understood by humans, but are useful for computational matching of
chemical entities. For rapid database lookups, the InChIKey is a
hashed key for the InChI, which has an invariant length of 14
characters. (Unfortunately, after implementation and publication of the hashing scheme it became apparent that in particular in case of stereochemically related molecules clashes of the InChI key take place in some, albeit rare cases.) Figure \ref{smiles} illustrates the SMILES, InChI and InChIKey for
lipoic acid.

@@

InChI or other identity-mapping algorithms allow for exact searching (within the
limitations of the algorithms). However, two other practically relevant algorithms for scientific
discovery based on chemical databases are \emph{substructure searching} and \emph{similarity
searching}, which are required to generalize from the data points contained in a particular database to other, related molecules. In substructure searching, the database is searched for a specified
wholly contained part of the structures in the database, while in similarity
searching, the database is searched for structures similar (in some property space) to a provided search, or query, 
structure. Chemical search packages are often implemented and optimized for a
given database technology; for example, the OrChem package is an open source
chemical search package for the Oracle database \cite{rijnbeek2009}.

Graph substructure matching is a variant of \emph{graph isomorphism detection},which is
widely believed to be computationally intractable [3]. To execute a graph isomorphism
search across a full chemical database of thousands or even millions of
structures is infeasible. For that reason, complex heuristics must be
used to drastically filter the candidate search space. Structure-based
\emph{fingerprints} are commonly used for this purpose. Fingerprints encode
characteristic features of a given chemical structure in a Boolean array, or
bitmap, and they can be implemented in a variety of ways. Common fingerprints are created by an algorithm which generates bitmap patterns for each atom in the structure; then each atom and its nearest
neighbors, including the bonds between them; then each group of atoms connected
by paths of increasing lengths up to the maximum implemented for the algorithm
(commonly around eight). In this way, characteristic parts of molecule are captured and the bitmap patterns are added together into the final fingerprint. If all bits in a query fingerprint are also present in the target
fingerprint of a stored database structure, only then is the structure subjected
to the computationally expensive subgraph matching algorithm. Bit operations are
very fast and independent of the number of atoms in a structure due to the fixed
length of the fingerprint.

Fingerprints also facilitate the rapid calculation of quantitative
structure-based similarity measures. An example measure of similarity is the
\emph{Tanimoto coefficient}, calculated as the ratio $T(a,b) = \frac{c}{(a+b-c)}$,
where $c$ is the count of bits on in the same position in both the two
fingerprints, $a$ is the count of bits on in the first structure, and $b$ is the
count of bits on in the second structure. Historically, this coefficient originated from the comparison of ecosystems around lakes, where a measure was needed to describe how similar those ecosystems are with respect to the animals, plants etc. they contain. However, it can be applied to every other system as well where features can be described in a binary manner. The Tanimoto coefficient varies in the
range 0.0 -- 1.0, with a score of 1.0 indicating that the two structures are either very
similar or even identical (i.e. their fingerprints are the same).

Structure-based similarity is however of
limited use in grouping of chemicals based on non-structural features, such as
shared bioactivity profiles. For this purpose, a large-scale and flexible
classification structure is needed for chemicals and their linked biological
information. One such answer to this need is provided in the form of \emph{ontologies}.
An ontology is a formal specification of entities and their relationships in a
domain of interest. It is structured around an underlying logical representation
such as Description Logics~\cite{baaderdl2007}. The logic-based representation allows
automated reasoning to derive new knowledge (inferences) from the knowledge
encoded, making the representation both compact and powerful.
Ontologies are already in widespread use across the biomedical domain, and in the cheminformatics field ontologies relating to the chemical structures themselves and the protein targets the molecules are active against are probably most common. A
similarity measure based on the information encoded in ontologies is termed
\emph{semantic similarity}, and such measures have already been applied to enhance
chemical classification in several problem areas \cite{couto2010}. Many open
questions still remain about the best way to combine the information encoded in
chemical structures with chemical ontologies \cite{hastingsowled2010}.

@@

Enumerating molecules is a combinatorial problem that has fascinated
chemists, computer scientists and mathematicians alike, for more than a
century. It is one important risk-reduction element in still being
able to encode closely-related molecules and for keeping the space
complexity of databases under (algorithmic) control. Indeed, while
trying to solve the problem of counting the isomers of paraffin
structures or counting substituted aromatic compounds, fundamental
principles in graph theory and combinatorics were developed by Cayley,
Polya and others. Even terms that are widely-used nowadays, like 'graph'
and 'tree', were originally coined in a chemistry context (Chapters 1 and
8 in \cite{faulon2010}). More than four decades ago, Joshua Lederberg,
Carl Djerassi, Edward Feigenbaum and others from Stanford University
started developing codes to explicitly enumerate molecules. Their
project, named DENDRAL, was funded by NASA with the goal of developing
an automated system to elucidate compounds from spectral data. While
studying this challenging problem, important concepts in computer
science were developed. DENDRAL is generally quoted in artificial
intelligence textbooks as a pioneer project that produced the first
expert system.

Enumerating molecules is not only an interesting academic exercise but has
practical utility as well. The foremost application of enumeration is \emph{structure
elucidation} of natural compounds such as metabolites. Metabolites are small molecules that can be found an organism (such as a human) and which in many cases indicate the state the organism is in - such as indicating a disease, or a fasted state, etc., and with a sufficiently large database the idea is to be able to detect the chemical composition of body fluids and to arrive at conclusions regarding the particular state of the organism at this point in time.
Ideally, the wishful bench chemist collects experimental data
from an unknown compound, the data is fed into a code, and the resulting
structure is returned. Although that streamlined process may not yield a unique
solution, there are commercial software products such as MOLGEN
(http://molgen.de) that can, for instance, list all structures matching a given
molecular formula, or mass spectrometry (MS) data. Another important application
of enumeration is \emph{molecular design}. Here the problem is to design compounds
(drugs, for instance) that optimize some physical, chemical, or biological
property or activity. Although less prolific and more recent than structure
elucidation, molecular design has introduced some novel solutions to molecular
enumeration. Finally, with the advent of combinatorial chemistry, molecular
enumeration has taken a central role as it allows computational chemists to
construct virtual libraries and test hypotheses, and it provides guidance in the
design of optimal combinatorial experiments.

The major difficulty with enumeration (apart from the size of the search space, which is thought to be in the order of 10^60 molecules for 30 heavy atoms) is that the \emph{in silico}
representation of molecules, the so-called molecular graphs, are
labeled objects (i.e. labeled graphs), while the atoms of a molecule
are of course not uniquely labeled. The mathematical concept to tackle
this problem is to consider orbits of labeled molecular graphs under
the operation of the symmetric group. However, when enumerating molecular
structures, one has thus to solve the so--called isomorphism problem, where molecules that seem different are actually equivalent (i.e., their atoms can be mapped on top of each other by keeping the connectivity between atoms intact).
The good news is that isomorphism can be solved in polynomial time for
molecular graphs due to the fact that molecules belong to a restricted
class of graphs known as \emph{bounded valence graphs}; however, that
same restriction prevents applying to a molecular graph the polynomial
delay enumeration results obtained for general graphs (Chapter 8 in
\cite{faulon2010}). While the complexity of molecular graph
enumeration remains an open problem it has nonetheless been shown that
molecules can be sampled efficiently \cite{goldberg1999}. Sampling
procedures based on the Metropolis or the Genetic Algorithms have been
developed and used to elucidate natural compounds from NMR data.

The underlying computational complexity and potential intractability
of molecular enumeration may not be a bottleneck after all, as
computational chemists have developed and successfully used
enumeration tools to generate large chemical libraries such as GDB-13,
comprising almost a trillion chemical
structures\footnote{http://www.dcb-server.unibe.ch/groups/reymond/gdb/home.html}. Yet,
the current enumeration software products do not generally produce
stereoisomers or tautomers; these require specific enumeration
procedures, which are still the subject of investigations by the
cheminformatics community. In addition, also GDB-13 only enumerates molecules up to 13 heavy atoms - given the possibly exponential increase in search space with the number of atoms, it seems unlikely that the practically relevant chemical space of about 30 heavy atoms will be exhaustively enumerated in the near future.

@@

The basis of predictive modeling in cheminformatics is that
\emph{chemical structure implies biological activity}. Thus, the goal
of any modeling approach is to capture and characterize the
correlations between structural features and the observed biological
activities. The goal is to describe cause-effect relationships with
the lowest possible risk, and also to describe the likelihood of error
(confidence) when using such models for decision making. In many
cases, the activity of a small molecule is due to its interaction with
a receptor. Traditional 'QSAR' approaches ignore receptor interactions,
focusing only on small molecule features, and therefore loose valuable
information. As a result, techniques such as proteochemometric
modeling have been designed to take into account both ligand and
receptor structures.

The first step in the predictive modeling of biological activities is
to describe compounds in a way amenable to computerized data mining, while at the same time capturing information relevant for the property one attempts to model. Here multiple approaches exist: Either we transform molecular graphs into a
numerical vector representation (a.k.a. \emph{descriptors} or
\emph{features}) of chemical structures, or we use algorithms for
comparing molecular graphs directly (\emph{molecule kernel}). In the
first case thousands of descriptors (many of which are equivalent)
have been described \cite{todeschini2000},
and a variety of tools are available for their calculation. Given a
set of molecules, observed activities and a descriptor vector, the
problem can be considered a traditional statistical modeling
problem. In the second case no explicit vector representation is
required, which prevents the challenge of transforming the molecular
graph into a vector without losing information. On the other hand, the
implicit kernel encoding imposes the challenge that we have to compare all compounds against
all compounds, which imposes large-scale mining challenge, which risks overrunning available
processing time and space. Note that the public PubChem dataset alone
contains over 30 million chemical compounds and over 70 million substances.
